Developing a Visual Studio Code Language Support Extension for the Snail Programming Language

### 2023-06-07 paper.tex
feat: adjust page margin
feat: move appendix after references
feat: draft directory structure section

\usepackage{cite}
\usepackage{hyperref}
\usepackage{listings}

@@

Much of software development in the present day takes place in integrated development environments (IDEs) \cite{JetBrains_2019}. An IDE is a collection of software development tools, such as a code editor, debugger, and build system, often unified under a similar user interface, with te goal of simplifying the software development process and enhancing developer productivity \cite{Gillis_Silverthorne_2018,Shyniaieva_2023}. In addition to compiling these tools together, many IDEs also offer advanced features within their code editors such as syntax highlighting, code auto-completion, and error-checking diagnostics. Today, there are any number of different IDEs available for use with any given programming language, many offered as freely-downloadable for public use. For example, a developer looking to write code in Java may choose to do so in Eclipse, IntelliJ IDEA, or BlueJ \cite{EclipseFoundation_2023, JetBrains_2023,KingsProgrammingEducationToolsGroup_2022}. With so many high quality IDEs available today, it is no surprise that 75\% of software developers today use an IDE in their everyday work \cite{JetBrains_2019}. Clearly, the IDE has become an integral part of the software development process today \cite{Vaniukov_2023}.

@@

The snail programming language (Strings Numbers Arrays and Inheritance Language) is a simple, object-oriented programming language meant to be implemented in a one-semester undergraduate course \cite{Angstadt_2023a}. In order to be implemented in a short time frame, snail is defined by limited features and a relatively annoying syntax. While this design makes snail easier to implement, it makes it hard for a develoepr to write programs in snail.

@@

This report presents the Snail Language Support VS Code extension, which seeks to address the lack of programming support tools for the snail programming language. Snail Langauge Support provides several important features to make programming in the snail language easier. First, it features syntax highlighting to make reading snail code easier and help highlight key structures or keywords of the language. Next, it features rudimentary auto completion with auto-closing brackets, braces, and quotes, as well as if-else, while loop, and class definition snippets, reducing the burden of memorizing snail's strict and unintuitive syntax. The extension also has automatic, real-time error checking diagnostics that allow a user to see syntax or parse errors in a snail program before running it for themselves. Finally, the extension is structured to support a full debugger with breakpoints, step-in, step-over, and step-out functionality.

@@

While VS Code is lightweight and fast, it still supports advanced features such as specific language "smarts", build tools, or debugging through the online VS Code extension marketplace \cite{Microsoft_2023b}. VS Code is designed for extensibility, providing robust APIs to allow independent developers to create and publish extensions \cite{Microsoft_2023d}. As a result, VS Code is able to grow and develop along with its community of users.

Currently popular extensions help make VS Code a popular code editor \cite{StackOverflow_2022}. The CodeSnap extension allows a user to capture and share improved screenshots of code \cite{adpyke_2021}. GitLens adds additional visualization tools to help developers contribute to Git repositories while editing in VS Code \cite{GitKraken_2023}. Language Support for Java(TM) provides intellisense, formatting, refactoring, and build system support for the Java language \cite{RedHat_2023}.

@@

@@

@@

@@

\bibliography{refs.bib}
\bibliographystyle{plain}
\end{document}

>>>>

\usepackage{cite}
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}
\usepackage{listings}

@@

Much of software development in the present day takes place in integrated development environments (IDEs) \cite{JetBrains_2019}. An IDE is a collection of software development tools, such as a code editor, debugger, and build system, often unified under a similar user interface, with te goal of simplifying the software development process and enhancing developer productivity \cite{Gillis_Silverthorne_2018,Shyniaieva_2023}. In addition to composing these tools together, many IDEs also offer advanced features within their code editors such as syntax highlighting, code auto-completion, and error-checking diagnostics. Today, there are any number of different IDEs available for use with any given programming language, many offered as freely-downloadable for public use. For example, a developer looking to write code in Java may choose to do so in Eclipse, IntelliJ IDEA, or BlueJ \cite{EclipseFoundation_2023, JetBrains_2023,KingsProgrammingEducationToolsGroup_2022}. With so many high quality IDEs available today, it is no surprise that 75\% of software developers today use an IDE in their everyday work \cite{JetBrains_2019}. Clearly, the IDE has become an integral part of the software development process today \cite{Vaniukov_2023}.

@@

The snail programming language (Strings Numbers Arrays and Inheritance Language) is a simple, object-oriented programming language meant to be implemented in a one-semester undergraduate course \cite{Angstadt_2023a}. In order to be implemented in a short time frame, snail is defined by limited features and a relatively annoying syntax. The language lacks a for-loop structure, opting instead to provide only while-loops. Each if statment \emph{requires} an else clause, even when the developer does not need to use one. Every statement must end with a semi-colon, which is not an issue until you accidentally forget one, and the resulting parse error message is wildly uninformative. While this design makes snail easier to implement, it makes it hard for a developer to write programs in snail.

@@

This report presents the Snail Language Support VS Code extension, which seeks to address the lack of programming support tools for the snail programming language. Snail Langauge Support provides several important features to make programming in the snail language easier. First, it features syntax highlighting to make reading snail code easier and help highlight key structures or keywords of the language. Further, it features rudimentary auto completion with auto-closing brackets, braces, and quotes, as well as if-else, while loop, and class definition snippets, reducing the burden of memorizing snail's strict and unintuitive syntax. The extension also has automatic, real-time error checking diagnostics that allow a user to see syntax or parse errors in a snail program before running it for themselves. Finally, the extension is structured to support a full debugger with breakpoints, step-in, step-over, and step-out functionality.

@@

While VS Code is lightweight and fast, it still supports advanced features such as build or debugger tools available through the online VS Code extension marketplace \cite{Microsoft_2023b}. VS Code is designed for extensibility, providing robust APIs to allow independent developers to create and publish extensions \cite{Microsoft_2023d}. As a result, VS Code is able to grow and develop along with its community of users.

Existing extensions make VS Code a popular code editor \cite{StackOverflow_2022}. The CodeSnap extension allows a user to capture and share improved screenshots of code \cite{adpyke_2021}. GitLens adds additional visualization tools to help developers contribute to Git repositories while editing in VS Code \cite{GitKraken_2023}. Language Support for Java(TM) provides intellisense, formatting, refactoring, and build system support for the Java language \cite{RedHat_2023}. The result of this extension support is an IDE that is used by over 75\% of professional and hobby software developers \cite{StackOverflow_2022}.

@@

TODO add signposting

@@

Most VS Code extensions share a similar directory structure. These structures may vary depending on what purpose the extension is meant to serve. In TODO FIGURE 1, we see a diagram of a general VS Code extension. First, the \lstinline{.vscode} directory configures how we run and test our exstension locally. Through the \lstinline{launch.json} and \lstinline{tasks.json} files, we define various tasks and configurations to streamline the build and testing process.\footnote{From personal experience, we strongly recommend taking the time to understand these files before diving into developing a VS Code extension. It will likely save painstaking debugging in the near future. See launch configurations: \url{https://code.visualstudio.com/docs/editor/debugging\#_launch-configurations} and tasks: \url{https://code.visualstudio.com/docs/editor/tasks}} It is important to note that these configuration files do not get packaged with a released VS Code extension. Next, a \lstinline{README.md} file that allows us to advertise the features of our extension. This file is rendered in the VS Code extension marketplace. We also have a \lstinline{tsconfig.json} file, which defines our TypeScript code gets transpiled to JavaScript, a topic we will discuss in further detail later in this paper.

The \lstinline{src} directory contains the source code that runs our extension. In this directory, the \lstinline{extension.ts} file defines special actions for our extension and launches the extension within an existing VS Code window. The \lstinline{server.ts} file defines a language server that allows our extension to take advantage of realtime error diagnostics. We further discuss language servers later in this paper.

Lastly, the \lstinline{package.json} file defines which features and configurations our extension supports. It also contains some biographical information about the extension, such as the author, publisher, or name of the exstension.

To build the structure for Snail Language Support, we used the \lstinline{Yeoman} tool to generate an extension skeleton \cite{Yeoman_2023}. This skeleton included configuration and source files for us to modify and build Snail Language Support on.

We build Snail Langauge Support with TypeScript. TypeScript is a syntactic superset of JavaScript that allows the developer the benefits of a type system without sacrificing JavaScript's flexibility \cite{TypeScript_2023}. Using the TypeScript compiler (\lstinline{tsc}), a program in TypeScript is transpiled to an equivalent JavaScript program. With a \lstinline{tsconfig.json} file, we can define how the TypeScript compiler handles our TypeScript code. We define where the compiler looks for our \lstinline{.ts} files, where to place resulting JavaScript files, and the strictness of type checks during transpilation.

@@

\bibliography{refs.bib}
\bibliographystyle{plain}

@@

\end{document}
